---
title: "RavenR: an open source R package to support flexible hydrologic modelling (code for figures)"
author: "R. Chlumsky"
date: "2021-09-27"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## `RavenR`: an open source R package to support flexible hydrologic modelling

This document provides the R scripts required to reproduce the figures in the use cases (Section 3) of the `RavenR` manuscript, submitted in 2021. All R code and model files can be sourced from [Github](https://github.com/rchlumsk/RavenR_manuscript_2021) and on Zenodo.

```{r load libraries}
library(ggplot2)
library(tidyr)
library(dplyr)
library(ggpubr)
library(xts)
library(tidyhydat)
library(weathercan)
library(DiagrammeR)
library(DiagrammeRsvg)
library(RavenR)
library(RColorBrewer)
```

### Download Liard Model and other files

```{r Download Liard River model files, message=FALSE, warning=FALSE, eval=FALSE}
## update to link to Github and/or Zenodo repository (https://github.com/rchlumsk/RavenR_manuscript_2021)

browseURL("https://github.com/rchlumsk/RavenR_manuscript_2021")

# download supporting files
download.file(url="https://raw.githubusercontent.com/rchlumsk/RavenR_manuscript_2021/main/Liard_unclean.rvh",
              destfile="Liard_unclean.rvh", mode="wb")
download.file(url="https://raw.githubusercontent.com/rchlumsk/RavenR_manuscript_2021/main/observation_stations.csv",
              destfile="observation_stations.csv", mode="wb")

# download model files zip file as raw
download.file(url="https://github.com/rchlumsk/RavenR_manuscript_2021/blob/main/LiardRiverModel.zip?raw=true",
              destfile="LiardRiverModel_test.zip")

# unzip LiardRiverModel.zip
if (file.exists(paste0(getwd(),"/","LiardRiverModel.zip"))) {
  # dir.create(paste0(outdir,"/","LiardRiverModel"))
  unzip(paste0(getwd(),"/","LiardRiverModel.zip"), exdir=paste0(getwd(),"/","LiardRiverModel"))
}

# create figures folder if it does not exist
if (!dir.exists("figures")) {
  dir.create("figures")
}
```

```{r download Raven and run Liard model if needed, message=FALSE, warning=FALSE, eval=FALSE}
# download Raven.exe from raven.uwaterloo.ca, store in RavenR extdata folder
## may require compilation of Raven.exe if not using Windows OS
rvn_download()

# run Raven Liard model, outputting to output_new for demonstration purposes
## produces same output as in the existing output folder
rvn_run(fileprefix="Liard", indir="./LiardRiverModel/",
        outdir="./LiardRiverModel/output_new/",
        showoutput=TRUE)
```


### Section 3.2 - Input file processing

#### Section 3.2.1 - Model configuration

```{r model configuration - rvi process plot}
d2 <- rvn_rvi_read("./LiardRiverModel/Liard.rvi") %>% 
  rvn_rvi_connections() %>% 
rvn_rvi_process_diagrammer(rvi_conn = ., lbl_size = 0.6)

# optionally, render graph
DiagrammeR::render_graph(d2)

d2 %>% 
  export_graph(file_name="./figures/rvi_process_diagrammer_Liard.pdf")
```

#### Section 3.2.2 - Forcing data

```{r forcing data - rvt met files}
# define coordaintes for station search (lat/long)
fort_liard <- c(60.241711, -123.467377)
stns <- weathercan::stations_search(coords=fort_liard, interval="day", 
                            dist = 500, starts_latest = 1985, ends_earliest = 2006)

# download 10 nearest stations for 1985-2005 inclusive
metdata <- weather_dl(stns$station_id[1:10], start="1985-10-01", end="2005-10-01", interval="day")

# interpolate missing values for the first 5 stations using data from all 10
## select for only the data columns we want to write to file - i.e. exclude non-interpolated data columns
metdata_interpolated <- metdata %>% 
  dplyr::select(station_name, station_id, lat, lon, elev, date, max_temp, min_temp, total_precip) %>%
  rvn_met_interpolate(cc=c("max_temp", "min_temp", "total_precip"), 
                      key_stn_ids = stns$station_id[1:5])
  
# write rvt files, including met_stndata.rvt
metdata_interpolated %>% 
  rvn_rvt_write_met()
```

#### Section 3.2.3 - Observation data

```{r observation data - rvt streamflow observation files}
# read in desired stations and their model subbasin IDs
obs_stns <- read.csv("observation_stations.csv")
  
# download daily flow data using tidyhydatgithub
hd <- tidyhydat::hy_daily_flows(station_number = obs_stns$stnID,
 start_date = "1985-01-01")

# Create RVT files - write locally with default names
rvn_rvt_tidyhydat(hd, subIDs=obs_stns$subID)
```

#### Section 3.2.3 - Model discretization file

```{r model discretization file - HRU cleaning for subbasin 3}

#Directory with uncleaned rvh
rvh_unclean <- rvn_rvh_read("Liard_unclean.rvh")
unclean_hrus <- rvh_unclean$HRUtable
unclean_subs <- rvh_unclean$SBtable

# number of uncleaned HRUs
nrow(unclean_hrus)

#Subset to subbasin 3
unclean_subs_3 <- subset(unclean_subs, unclean_subs$SBID==3)
unclean_hrus_3 <- subset(unclean_hrus, unclean_hrus$SBID==3)

# number just in subbasin 3
nrow(unclean_hrus_3)

# check all hrus
unclean_hrus_3$LandUse %>% unique()

#Set protected hrus to any hru which is a wetland OR water type
protected_hrus <- unclean_hrus_3$ID[(unclean_hrus_3$LandUse=="WETLAND" | 
                                     unclean_hrus_3$LandUse=="WATER")]

# Set locked hrus to glacier type
locked_hrus <- unclean_hrus_3$ID[(unclean_hrus_3$LandUse=="GLACIER")]

#Clean HRUs based on area tolerance equal to 0.5%
#Reduces number of HRUs in subbasin 3 from 172 to 87
clean_hrus_v1 <- rvn_rvh_cleanhrus(HRUtab=unclean_hrus_3, 
                                   SBtab=unclean_subs_3, 
                                   area_tol = 0.005, 
                                   ProtectedHRUs = protected_hrus,
                                   LockedHRUs = locked_hrus,
                                   merge = TRUE)

#Increase area tolerance to 2%
#Reduces number of HRUs in subbasin 3 from 172 to 44
clean_hrus_v2 <- rvn_rvh_cleanhrus(HRUtab=unclean_hrus_3, 
                                   SBtab=unclean_subs_3, 
                                   area_tol = 0.02, 
                                   ProtectedHRUs = protected_hrus,
                                   LockedHRUs = locked_hrus,
                                   merge = TRUE)

# check total areas
unclean_hrus_3$Area %>% sum()
clean_hrus_v1$Area %>% sum()
clean_hrus_v2$Area %>% sum()

#Create Plots to Show how distribution of land cover, elevation , aspect and slope changes
unclean_hrus_3$Type <- "Uncleaned"
clean_hrus_v1$Type <- "Cleaned 0.5% Area Tolerance"
clean_hrus_v2$Type <- "Cleaned 2.0% Area Tolerance"

all_hrus <- rbind(unclean_hrus_3, clean_hrus_v1, clean_hrus_v2)
all_hrus$Type <- factor(all_hrus$Type, levels = c("Uncleaned","Cleaned 0.5% Area Tolerance","Cleaned 2.0% Area Tolerance"))

## checks changes in land use numerically
all_hrus %>% 
  group_by(LandUse,Type) %>% 
  summarise(Area=sum(Area))

# Plot distribution of land class
p1 <- all_hrus %>% 
  ggplot()+
  geom_bar(aes(x=LandUse, y=Area, fill = Type), stat = "summary", position = "dodge", fun="sum")+
  scale_fill_grey(name="Clean Type", labels=c("Uncleaned","0.5% Area\nTolerance", "2.0% Area\nTolerance"))+
  labs(x = 'Land Class', y = expression(paste('Area (', km^2, ')')))+
  theme_bw()+
  theme(legend.key.size = unit(1.2, "cm"))+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
  
p1 

p1 %>% 
  ggsave(filename="./figures/rvh_cleanhrus_Liard.pdf",
       plot=., units="in",width=6,height=4)
```


### Section 3.3 - Output file processing and analysis

### Section 3.3.1 - Evaluation of model realism

```{r evaluation of model realism}

## Budyko plot - plot 1 ----

# add :WriteForcingFunctions and :WriteMassBalanceFile to rvi, plot buydko curve
# note: can alternatively get AET, PET, PRECIP by using custom output
# AET_Daily_Average_ByWatershed;

wstor <- rvn_watershed_read("./LiardRiverModel/output/WatershedStorage.csv")
ff <- rvn_forcings_read("./LiardRiverModel/output/ForcingFunctions.csv")

precip <- ff$forcings$rain+ff$forcings$snow
pet <- ff$forcings$PET
aet <- diff.xts(x=wstor$watershed_storage$Cum..Losses.to.Atmosphere..mm.,
                k=1, na.pad=TRUE)
aet[1] <- 0

x <- merge.xts(precip,pet,aet)
names(x) <- c("precip","pet","aet")

p1 <- rvn_budyko_plot(x=x, budyko_curve = TRUE, limiting_labels = FALSE) +
  annotate("text",x=-Inf,y=Inf,label="(A)",fontface="bold",size=6, hjust=-0.1,vjust=1.2)

p1


## regime curve plot 2 ----

rvh <- rvn_rvh_read("./LiardRiverModel/Liard.rvh")
Liard_DA <- rvh$SBtable[rvh$SBtable$SBID == 63, ]$TotalUpstreamArea * 1e6 # m2

qq <- rvn_hyd_read("./LiardRiverModel/output/Hydrographs.csv") %>% 
  rvn_hyd_extract(subs="SUB_63", hyd=.) 

# convert m3/s to mm/d
qq_mmd <- qq*36400*1000/Liard_DA

# merge and process
regime_xx <- merge.xts(qq_mmd, x)

regime_monthly_df <- apply.monthly(regime_xx, mean, na.rm=TRUE) %>% 
  rvn_fortify_xts() %>% 
  mutate(Month = lubridate::month(Date), Year = lubridate ::year(Date)) %>% 
  group_by(Month) %>% 
  summarise_all(mean)

p2 <- regime_monthly_df %>% 
  select(Month,sim,obs,precip,pet,aet) %>% 
  pivot_longer(cols=-Month, names_to="regime_type",values_to="regime_value") %>% 
ggplot(data=., aes(x=Month, y=regime_value, color=regime_type))+
  geom_line(lwd=1.2)+
  rvn_theme_RavenR()+
  ylab("Depth (mm)")+
  scale_x_continuous(breaks=seq(1,12),
                     expand=c(0,0))+
  theme(legend.position = c(0.86, 0.73),
        legend.background = element_rect(fill = "white", color = "black"))+
  scale_colour_manual(values=brewer.pal(5,"Set2"))+
  # annotate("text",x=2,y=3.2,label="(B)",fontface="bold",size=6, hjust=0)
  annotate("text",x=-Inf,y=Inf,label="(B)",fontface="bold",size=6, hjust=-0.1,vjust=1.2)+
  theme(legend.key.size = unit(0.5, 'cm'))

p2
  

## soil storage stationarity checks - plot 3 ----

prd <- "1987-10-01/1996-06-01"

# note: can also use the CustomOutput to get soil moisture outputs
wstor <- rvn_watershed_read("./LiardRiverModel/output/WatershedStorage.csv")

# change parameter in just one subbasin and check
p3 <- rvn_fortify_xts(wstor$watershed_storage[prd]) %>% 
  select(Date,starts_with("Soil")) %>% 
  pivot_longer(cols=starts_with("Soil"), 
               names_to="soil_number",
               values_to="soil_storage") %>% 
  ggplot()+
  geom_line(aes(x=Date, y=soil_storage, linetype=soil_number, color=soil_number))+
  ylab("Soil Storage (mm)")+
  rvn_theme_RavenR()+
  # scale_colour_discrete(name="Legend") +
  scale_linetype_manual(name="Soil Layer", values=c(1,2,1,2),
                        labels=as.character(seq(1,4))) +
  scale_color_manual(name="Soil Layer", values=brewer.pal(n=4, "Set2"),
                     labels=as.character(seq(1,4)))+
  # theme()+
  guides(color = guide_legend(override.aes = list(size = 1)))+
  theme(legend.background = element_rect(fill = "white", color = "black"),
        legend.direction = "vertical",
        legend.box = "horizontal",
        legend.position = c(0.79,0.975), # c(0.025,0.975)
        legend.justification = c(0, 1),
        legend.key.size=unit(0.12,"in"))+
  theme(legend.title=element_text(size=8))+
  guides(color = guide_legend(override.aes = list(size = 1)))+
  annotate("text",x=as.Date("1987-10-01"),y=60,label="(C)",fontface="bold",size=6, hjust=0.4, vjust=-0.7)

p3


## snow plots - plot 4 ---- 

# read in snow custom output
snow <- rvn_custom_read("./LiardRiverModel/output/SNOW_Daily_Average_ByHRU.csv", no_runname = TRUE)
names(snow) <- sprintf("HRU_%s",names(snow))

# read in observed snow
snow_obs <- read.csv("./LiardRiverModel/data_obs/Snow/Frances River.csv")
snow_obs$Date <- as.Date(snow_obs$Date, format = "%m/%d/%Y")
snow_obs_xts <- xts(snow_obs$SWE, order.by=snow_obs$Date)
names(snow_obs_xts) = c("snow_obs")

# create plot
p4 <-
  snow$HRU_819[prd] %>% 
  rvn_fortify_xts() %>% 
  ggplot()+
  geom_line(aes(x=Date, y=HRU_819, colour="Simulated", linetype="Simulated"))+
  geom_point(aes(x=Date, y=HRU_819, colour="Simulated", shape="Simulated"))+
  geom_point(data=rvn_fortify_xts(snow_obs_xts[prd]),
                aes(x=Date,y=snow_obs,colour="Observed", shape="Observed"))+
  scale_color_manual(name="Legend",values=c("Simulated"="darkgrey","Observed"="black"))+
  scale_linetype_manual(name="Legend",values=c("Simulated"="solid","Observed"="blank"))+
  scale_shape_manual(name="Legend",values=c("Simulated"=NA,"Observed"=19))+
  ylab("Snow SWE (mm)")+
  rvn_theme_RavenR()+
  theme(legend.background = element_rect(fill = "white", color = "black"),
        legend.direction = "vertical",
        legend.box = "horizontal",
        legend.position = c(0.7,0.975), # c(0.025,0.975)
        legend.justification = c(0, 1),
        legend.key.size=unit(0.15,"in"))+
  annotate("text",x=as.Date("1988-01-01"),y=300,label="(D)",fontface="bold",size=6, hjust=0.7, vjust=-0.0)

p4

# print party figure realism plots to file
list(p1,p2,p3,p4) %>% 
ggarrange(plotlist=., ncol=2, nrow=2, common.legend=FALSE,
          align='hv') %>% 
ggsave(filename="./figures/mixed_custom_checks_Liard.pdf",
       plot=.,
         units="in",
          width=8,height=6)
```

### Section 3.3.2 - Evaluation of model performance

```{r evaluation of model performance}

hyd <- rvn_hyd_read("./LiardRiverModel/output/Hydrographs.csv")
sub63 <- rvn_hyd_extract(subs="SUB_63",hyd = hyd)

p1 <- rvn_hyd_plot(sim=sub63$sim, obs=sub63$obs,
             prd="1989-03-01/1992-12-01",
             winter_shading = FALSE)+
  scale_colour_manual(name='Legend',values=c("Obs"='black',"Sim"='darkorange'))+
  annotate("text",x=as.Date("1989-02-01"),y=14000,label="(A)",fontface="bold",size=6, hjust=0.3, vjust=0.2)

p1

p2 <- rvn_annual_peak(sim=sub63$sim, obs=sub63$obs)$p1+
  theme(legend.position = 'none')+
  annotate("text",x=-Inf,y=Inf,label="(B)",fontface="bold",size=6, hjust=-0.1,vjust=1.2)

p2

# plot with axis labels reduced for p3
xx <- seq(1986,2005)
xx_lbl <- xx
xx_lbl[which(xx %in% (seq(1986,2005,by=2)+1))] <- NA
xx_lbl <- as.character(xx_lbl)
xx_lbl[which(is.na(xx_lbl))] <- ""

p3 <- rvn_annual_peak_timing_error(sim=sub63$sim, obs=sub63$obs, add_labels = TRUE)$p1+
  theme(legend.position = 'none')+
  annotate("text",x=-Inf,y=Inf,label="(C)",fontface="bold",size=6, hjust=-0.1,vjust=1.2)+
  scale_x_discrete(breaks=xx,
        labels=as.character(xx_lbl))

p3

p4 <- rvn_cum_plot_flow(sim=sub63$sim["1989-03-01/1992-12-01"], obs=sub63$obs["1989-03-01/1992-12-01"])+
  scale_colour_manual(values=c('black','darkorange')) +
  theme(legend.position = 'none')+
  # annotate("text",x=-Inf,y=Inf,label="(E)",fontface="bold",size=6, hjust=-0.1,vjust=1.2)
  annotate("text",x=as.POSIXct("1989-04-01"),y=1e11,label="(D)",fontface="bold",size=6, hjust=0.5, vjust=0.2)

p4

# print party figure diagnostic plots to file
list(p1,p2,p3,p4) %>% 
ggarrange(plotlist=., ncol=2, nrow=2, common.legend=TRUE,
          align='hv', legend='bottom') %>% 
ggsave(filename="./figures/mixed_diagnostic_plots_Liard.pdf",
       plot=.,
         units="in",
          width=8,height=6)
```
